### 2024-09-15 manuscript/manuscript.qmd
add flowchart to methods

- Numerous studies over the last 25 years have established what emotions listeners perceive and recognise in music [@gomez2021]. In the last 15 years, it has become possible to trace the recognised emotions to musical contents such as expressive features [@lindstrom2003expressivity], structural aspects of music [@eerola_friberg_bresin_2013;@anderson2022ex;@grimaud_eerola_2022], or acoustic features [@yang2008;@panda2013multi;@saari_et_al_2015;@eerola2011c] or emergent properties identified through deep learning [@er2019music;@sarkar2020recognition].

@@

We preregistered our meta-analysis plan on 21 June 2024 at OSF, [https://osf.io/c5wgd](https://osf.io/c5wgd), and the plan is also available at [https://tuomaseerola.github.io/metaMER/preregistration/preregistration.html](https://tuomaseerola.github.io/metaMER/preregistration/preregistration.html)).

In the search stage, we used three databases, _Web of Science_, _Scopus_, and _Open Alex_ to identify journal articles published between 2014 and 2024 containing keywords/title `valence OR arousal OR classi OR categor OR algorithm AND music 
 AND emotion AND recognition` (see specific search strings for each database in [SI](http://)). All searches were done in May 2024.

The initial search yielded 553 potential studies, which were interactively screened for relevance in three stages by both authors, resulting in 46 studies that passed our inclusion criteria (define here: has included some form of emotion modelling with a set of features, and outcome measures such as $R^2$ or classification accuracy ($F1$ score or _precision_ or _recall_) and overall quality control; see [SI](http://) for a breakdown). After the screening stage, we define a set of entities to extract characterising music (genre, stimulus number [N], duration), features extracted (number, type, source, defined by [@panda2020audio]), model type (regression, neural network, SVM, etc.) and outcome measure ($R^2$, _MSE_, _MCC_), model complexity (??? model architecture?), and type of model cross-validation. We converted all regression results into $R^2$ values for valence and arousal and classification results into Matthews correlation coefficient [_MCC_, @chicco2020advantages].

[optional Figure: flowchart of the study inclusions/eliminations]

@@

@@

- Decision: Report all models or the best model?
  * Possible solution: Report all models, trimmed models, and but mainly rely on Max

@@

| Concept     | Models, obs|$r$ [95%-CI]            | $t$ | $p$  |$\tau^2$|$I^2$| Prediction interval
|:------------|:----------|:--------  --------------|:----|:-----|:-----|:------|
| Valence All | 49,42480  | 0.6137 [0.5540; 0.6672] | 15.9|.0001 | 0.092| 98.1% |
| Valence Trim| 27,18402  | 0.6048 [0.5920; 0.6172] | 72.4|.0001 | 0.0007| 33.1%|
| Valence M   | 14,10641  | 0.6038 [0.4671; 0.7124] | 7.83|.0001 | 0.100| 97.8% |
| Valence Md  | 14,10641  | 0.6018 [0.4659; 0.7099] | 7.87|.0001 | 0.098| 97.7% |
| Valence Max | 14,10641  | 0.6520 [0.5026; 0.7636] | 7.44|.0001 | 0.142| 97.9% | -0.0719; 0.9260
| Val. MaxTrim|   8,7205  | 0.6140 [0.5538; 0.6679] | 18.5|.0001 | 0.008| 88.5% | 0.4470; 0.7396
| *N Features*|           |                         |     |      |      |       |
| <18 F       |           | 0.xxxx [0.xxxx; 0.xxxx] | xxxx|.xxxx | 0.xxx| XXXX% |
| 18-260 F    |           | 0.xxxx [0.xxxx; 0.xxxx] | xxxx|.xxxx | 0.xxx| XXXX% |
| 260+ F      |           | 0.xxxx [0.xxxx; 0.xxxx] | xxxx|.xxxx | 0.xxx| XXXX% |
| *Techniques*|           |                         |     |      |      |       |
|   KS        |           | 0.3606 [0.3201; 0.3997] | xxxx|.xxxx | 0.xxx| XXXX% |
|   LM        |           | 0.xxxx [0.xxxx; 0.xxxx] | xxxx|.xxxx | 0.xxx| XXXX% |
|   FD        |           | 0.xxxx [0.xxxx; 0.xxxx] | xxxx|.xxxx | 0.xxx| XXXX% |
|   NN        |           | 0.xxxx [0.xxxx; 0.xxxx] | xxxx|.xxxx | 0.xxx| XXXX% |
|   RF        |           | 0.xxxx [0.xxxx; 0.xxxx] | xxxx|.xxxx | 0.xxx| XXXX% |

@@

see this to ignore I^2 and rely on prediction interval: 
https://onlinelibrary.wiley.com/doi/full/10.1002/jrsm.1678

@@

| Concept     | Models, obs|$r$ [95%-CI]             | $t$ | $p$  |$\tau^2$|$I^2$|
|:------------|:-----------|:------------------------|:----|:-----|:-----|:------|
| Arousal     | 55,43160   | 0.7959 [0.7666; 0.8218] |29.0 |0.0001|0.0676|95.6%  |
| Arousal Trim| 27,18402   | 0.7932 [0.7846; 0.8014] | 96.5|0.0001|0.0030|73.3%  |
| Arousal M   | 14,10641   | 0.7567 [0.6752; 0.8199] | 12.7|0.0001|0.0739|95.6%  |
| Arousal Md  | 14,10641   | 0.7627 [0.6819; 0.8252] | 12.7|0.0001|0.0757|95.3%  |
| Arousal Max | 14,10641   | 0.8077 [0.7088; 0.8755] | 10.3|0.0001|0.155 | 96.8%  | 0.2258; 0.9648
| Aro.Max.Trm | 11,10021   | 0.7869 [0.7490; 0.8197] | 25.6|0.0001|0.0132| 96.8%  | 0.6372; 0.9047

| *N Features*|            |          |              |     |      |      |       |   
| <18 F       |            |          |              |     |      |      |       |
| 18-260 F    |            |          |              |     |      |      |       |
| 260+ F      |            |          |              |     |      |      |       |
| *Techniques*|            |          |              |     |      |      |       |  
|   KS        |            |          |              |     |      |      |       |
|   LM        |            |          |              |     |      |      |       |
|   FD        |            |          |              |     |      |      |       |
|   NN        |            |          |              |     |      |      |       |
|   RF        |            |          |              |     |      |      |       |

@@

| Model       | Models, obs|$r$ [95%-CI]             | $t$ | $p$  |$\tau^2$|$I^2$|
|:------------|:-----------|:------------------------|:----|:-----|:-----|:------|
| All         | 89,87347   | 0.8074 [0.7681; 0.8407] |21.4 |0.0001|0.2415|99.7%  |
| All Trim    | 29,6499    | 0.8185 [0.8046; 0.8314] | 58.3|0.0001|0.0066| 60.4%  |
| Max         | 14,17184   | 0.8564 [0.7386; 0.9234] | 10.3|0.0001|0.329 | 99.8% |
| Max Trim    | 6,3653     | 0.8689 [0.7760; 0.9249] | 11.6|0.0001|0.0749| 97.5% |

@@

## Main outcomes 
- Arousal is easier to predict (r = 0.7627) than valence (r = 0.6236), as we predicted. The glass ceiling seems to be at ...
- Classification ...
- Model accuracy is surprisingly little affected by the number of features (?) or modelling technique (?). 
- Some of the complex state-of-the-art techniques (e.g., NNs) do not deliver impressive improvements over older techniques (e.g., SVR, RF)
- Variation in study/model/data quality is large and can be seen in heterogenuity and the amount of studies eliminated

## Calls for action/points to improve in such studies
- *Documentation* the details in full (features, stimuli, model details, cross-validation)
- *Quality* of the underlying data (emotion ratings, classes, or even stimulus properties?
- *Generalisibility* of the models (some studies such as X and Y address this by applying the models across several datasets)
- Diversity in the evaluative aspects of studies: *overfitting*, numerous ways of cross-validating, not sharing data or analysis scripts, not reporting in the same way
- What proportion of stimuli are Western music, and what genres tend to dominate?

@@

Study preregistration, data, analysis scripts and supporting information is available at Github, [https://tuomaseerola.github.io/metaMER](https://tuomaseerola.github.io/metaMER).

>>>>

-   Numerous studies over the last 25 years have established what emotions listeners perceive and recognise in music [@gomez2021]. In the last 15 years, it has become possible to trace the recognised emotions to musical contents such as expressive features [@lindstrom2003expressivity], structural aspects of music [@eerola_friberg_bresin_2013; @anderson2022ex; @grimaud_eerola_2022], or acoustic features [@yang2008; @panda2013multi; @saari_et_al_2015; @eerola2011c] or emergent properties identified through deep learning [@er2019music; @sarkar2020recognition].

@@

We preregistered the meta-analysis plan on 21 June 2024 at OSF, <https://osf.io/c5wgd>, and the plan is also available at <https://tuomaseerola.github.io/metaMER/preregistration/preregistration.html>).

In the search stage, we used three databases, *Web of Science*, *Scopus*, and *Open Alex* to identify journal articles published between 2014 and 2024 containing keywords/title `valence OR arousal OR classi* OR categor* OR algorithm AND music AND emotion AND recognition` (see specific search strings for each database in [SI](http://)). All searches were done in May 2024.

The initial search yielded 553 potential studies, which were interactively screened for relevance in three stages by both authors, resulting in 46 studies that passed our inclusion criteria (define here: has included some form of emotion modelling with a set of features, and outcome measures such as $R^2$ or classification accuracy ($F1$ score or *precision* or *recall*) and overall quality control; see [SI](http://) for a breakdown). After the screening stage, we define a set of entities to extract characterising music (genre, stimulus number \[N\], duration), features extracted (number, type, source, defined by [@panda2020audio]), model type (regression, neural network, SVM, etc.) and outcome measure ($R^2$, *MSE*, *MCC*), model complexity (??? model architecture?), and type of model cross-validation. We converted all regression results into $R^2$ values for valence and arousal and classification results into Matthews correlation coefficient [*MCC*, @chicco2020advantages].

\[Figure 1: flowchart of the study inclusions/eliminations\]

```{r echo=FALSE}
library(glue)
library(DiagrammeR)

a_label <- glue('Records identified from:
                Web of Science (n = 142)
                Scopus (n = 227)
                OpenALEX (n = 278)')
b_label <- glue('Records removed before screening: 
                Duplicates entries/errors (n = 94)')
e_label <- glue('Records removed: 
                Irrelevant titles/themes (n = 338)
                Abstract screening (n = 95)
                Discussion (n = 24)')
h_label <- glue('Full text articles excluded
             (n = 50)')
grViz("
digraph cohort_flow_chart
{
node [fontname = Helvetica, fontsize = 12, shape = box, width = 4]
a[label = '@@1']
b[label = '@@2']
c[label = 'Records after duplicates removed (n = 553)']
d[label = 'Records screened (n = 553)']
e[label = '@@3']
f[label = 'Full text articles assessed (n = 96)']
g[label = 'Studies included (n = 46)']
h[label = '@@4']

{ rank = same; a b}
{ rank = same; d, e}
{ rank = same; f, h}

a -> b;
a -> c;
c -> d;
d -> e [ minlen = 3 ];
d -> f;
f -> h [ minlen = 3 ];
f -> g;
}

[1]: a_label
[2]: b_label
[3]: e_label
[4]: h_label

")
```

@@

|            | Regression                 | Classification            |
|:-----------|:---------------------------|:--------------------------|
| Study N    | 24                         | 14                        |
| Model N    | 258                        | 108                       |
| Techniques | Flexible Discriminants: 64 | 37                        |
| Techniques | KS, Add. & KNN: 24         | 8                         |
| Techniques | Linear Methods: 74         | 24                        |
| Techniques | Neural Nets: 74            | 27                        |
| Techniques | Random Forests: 22         | 12                        |
| Feature N  | Min=3, Md=472.5, Max=654   | Min=3, Md=231, Max=8904   |
| Stimulus N | Min=40, Md=324, Max=2486   | Min=124, Md=387, Max=5192 |

@@

-   Decision: Report all models or the best model?

    -   Possible solution: Report all models, trimmed models, and but mainly rely on Max

@@

| Concept      | Models, obs | $r$ \[95%-CI\]            | $t$  | $p$   | $\tau^2$ | $I^2$ |
|:----------|:----------|:----------|:----------|:----------|:----------|:----------|
| Valence All  |             | 0.6137 \[0.5540; 0.6672\] | 15.9 | .0001 | 0.092    | 98.1% |
| Valence Trim |             | 0.6048 \[0.5920; 0.6172\] | 72.4 | .0001 | 0.0007   | 33.1% |
| Valence M    |             | 0.6038 \[0.4671; 0.7124\] | 7.83 | .0001 | 0.100    | 97.8% |
| Valence Md   |             | 0.6018 \[0.4659; 0.7099\] | 7.87 | .0001 | 0.098    | 97.7% |
| Valence Max  | 24,15660    | 0.6585 \[0.5574; 0.7404\] | 7.44 | .0001 | 0.142    | 97.9% |
| Val. MaxTrim | 13,7363     | 0.6489 \[0.5980; 0.6945\] | 18.5 | .0001 | 0.008    | 88.5% |
| *N Features* |             |                           |      |       |          |       |
| \<18 F       |             | 0.xxxx \[0.xxxx; 0.xxxx\] | xxxx | .xxxx | 0.xxx    | XXXX% |
| 18-260 F     |             | 0.xxxx \[0.xxxx; 0.xxxx\] | xxxx | .xxxx | 0.xxx    | XXXX% |
| 260+ F       |             | 0.xxxx \[0.xxxx; 0.xxxx\] | xxxx | .xxxx | 0.xxx    | XXXX% |
| *Techniques* |             |                           |      |       |          |       |
| KS           |             | 0.3606 \[0.3201; 0.3997\] | xxxx | .xxxx | 0.xxx    | XXXX% |
| LM           |             | 0.xxxx \[0.xxxx; 0.xxxx\] | xxxx | .xxxx | 0.xxx    | XXXX% |
| FD           |             | 0.xxxx \[0.xxxx; 0.xxxx\] | xxxx | .xxxx | 0.xxx    | XXXX% |
| NN           |             | 0.xxxx \[0.xxxx; 0.xxxx\] | xxxx | .xxxx | 0.xxx    | XXXX% |
| RF           |             | 0.xxxx \[0.xxxx; 0.xxxx\] | xxxx | .xxxx | 0.xxx    | XXXX% |

@@

see this to ignore I\^2 and rely on prediction interval: https://onlinelibrary.wiley.com/doi/full/10.1002/jrsm.1678

@@

| Concept      | Models, obs | $r$ \[95%-CI\]            | $t$  | $p$    | $\tau^2$ | $I^2$ |
|:----------|:----------|:----------|:----------|:----------|:----------|:----------|
| Arousal      |             | 0.7959 \[0.7666; 0.8218\] | 29.0 | 0.0001 | 0.0676   | 95.6% |
| Arousal Trim |             | 0.7932 \[0.7846; 0.8014\] | 96.5 | 0.0001 | 0.0030   | 73.3% |
| Arousal M    |             | 0.7567 \[0.6752; 0.8199\] | 12.7 | 0.0001 | 0.0739   | 95.6% |
| Arousal Md   |             | 0.7627 \[0.6819; 0.8252\] | 12.7 | 0.0001 | 0.0757   | 95.3% |
| Arousal Max  | 24,15660    | 0.8070 \[0.7453; 0.8550\] | 10.3 | 0.0001 | 0.155    | 96.8% |
| Aro.Max.Trm  | 14,12061    | 0.8182 \[0.8021; 0.8331\] | 25.6 | 0.0001 | 0.0132   | 96.8% |
| *N Features* |             |                           |      |        |          |       |
| \<18 F       |             |                           |      |        |          |       |
| 18-260 F     |             |                           |      |        |          |       |
| 260+ F       |             |                           |      |        |          |       |
| *Techniques* |             |                           |      |        |          |       |
| KS           |             |                           |      |        |          |       |
| LM           |             |                           |      |        |          |       |
| FD           |             |                           |      |        |          |       |
| NN           |             |                           |      |        |          |       |
| RF           |             |                           |      |        |          |       |

@@

| Model    | Models, obs | $r$ \[95%-CI\]            | $t$  | $p$    | $\tau^2$ | $I^2$ |
|:----------|:----------|:----------|:----------|:----------|:----------|:----------|
| All      | 89,87347    | 0.8074 \[0.7681; 0.8407\] | 21.4 | 0.0001 | 0.2415   | 99.7% |
| All Trim | 29,6499     | 0.8185 \[0.8046; 0.8314\] | 58.3 | 0.0001 | 0.0066   | 60.4% |
| Max      | 14,17184    | 0.8564 \[0.7386; 0.9234\] | 8.32 | 0.0001 | 0.329    | 99.8% |
| Max Trim | 6,3653      | 0.8689 \[0.7760; 0.9249\] | 11.6 | 0.0001 | 0.0749   | 97.5% |

@@

## Main outcomes

-   Arousal is easier to predict (r = 0.7627) than valence (r = 0.6236), as we predicted. The glass ceiling seems to be at ...
-   Classification ...
-   Model accuracy is surprisingly little affected by the number of features (?) or modelling technique (?).
-   Some of the complex state-of-the-art techniques (e.g., NNs) do not deliver impressive improvements over older techniques (e.g., SVR, RF)
-   Variation in study/model/data quality is large and can be seen in heterogenuity and the amount of studies eliminated

## Calls for action/points to improve in such studies

-   *Documentation* the details in full (features, stimuli, model details, cross-validation)
-   *Quality* of the underlying data (emotion ratings, classes, or even stimulus properties?
-   *Generalisibility* of the models (some studies such as X and Y address this by applying the models across several datasets)
-   Diversity in the evaluative aspects of studies: *overfitting*, numerous ways of cross-validating, not sharing data or analysis scripts, not reporting in the same way
-   What proportion of stimuli are Western music, and what genres tend to dominate?

@@

Study preregistration, data, analysis scripts and supporting information is available at Github, <https://tuomaseerola.github.io/metaMER>.
